The role of aneuploidy in the evolution of cancer drug resistance

### 2024-02-04 EvolutionaryRescue1.tex
continue editing results

We can categorize tumors by their size: small tumors with size $N<N_a^*$, intermediate tumors with size $N_a^*<N<N_m^*$, and large tumors with size $N>N_m^*$ .
For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_w=0.14, u=10^{-2}, v=10^{-7}$, we are in the partially resistant aneuploid case, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4.4 \cdot 10^6$ and $N_m^* \approx 4 \cdot 10^7$.
Hence, we obtain the ratio $N^*_a/N^*_m \approx 0.11$ (\cref{eq:N_ratio}), that is, aneuploidy reduces the threshold population size by approximately 90\% 
Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4.19 \cdot 10^6$ cells for a wide variety of tumors as reported by \citet{avanzini2019cancer}.

@@

Therefore, the ratio of threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the ratio with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost, $s$, such that 
\begin{equation}\label{eq:Nsgv}
\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{s}{\abs{\Delta_w}}.
\end{equation}
Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_denovo}).  % TODO which approximations? unclear
Standing genetic variation will drive evolutionary rescue to the drug if wildtype growth rate $\Delta_w$ is very negative due to a stronger effect of the drug on sensitive cells, or if the aneuploidy cost $s$ is very small due to a low cost of aneuploidy in the pre-drug conditions.
In contrast, de-novo aneuploid cells will have a greater contribution to rescue if the aneuploidy cost, $s$, is large, the effect of the drug is weak ($\Delta_w$ is small), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
For example, with  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=10^{-3}, s=7.3\cdot10^{-3}$, the ratio of the threshold population sizes for standing vs de-novo variation is $\tilde{N}^*_a/N^*_a \approx 1.825$, which means that de-novo genetic variation is  the main driver of evolutionary rescue.

Using \cref{eq:N_a,eq:N_m,eq:Nsgv}, we can find the ratio of threshold tumor size for rescue via standing genetic variation  to the threshold for direct mutation,
\begin{equation} \label{eq:N_sgv_ratio}
\frac{\tilde{N}^*_a}{N^*_m}=\frac{\tilde{N}^*_{a}}{N^*_{a}} \cdot \frac{N^*_a}{N^*_m} \approx \frac{s}{\abs{\Delta_w}}\begin{cases}
    \frac{\abs{\Delta_a}}{\tilde{u}\lambda_a} ,&
  \Delta_a T^* \ll -1 ,\\ 
  \frac{1}{\tilde{u}}\left(v \cdot \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1  ,\\ 
  v \frac{\Delta_m}{\lambda_m} \cdot \left(\tilde{u}\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1 .
  \end{cases}
\end{equation}
We observe that evolutionary rescue through direct mutation is more likely if the cost of aneuploidy $s$ is very large or the effect of the drug $\Delta_w$ is small.  In contrast, standing genetic variation will drive adaptation if the pre-treatment chromosome missegreagation rate $\tilde{u}$ is very large. The ratio does not depend on the rate of chromosome missegregation induced by the drug $u$. However, if the growth rate of aneuploidy $\Delta_a$ increases then evolutionary rescue is driven by standing genetic variation. For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, \tilde{u}=10^{-3}, v=10^{-7}$ we are in the partially resistant aneuploid case and obtain the ratio $\tilde{N}^*_a/N^*_m\approx\frac{s}{\abs{\Delta_w}}\times\frac{\abs{\Delta_a}}{\tilde{u}\lambda_a}=0.356$ which means that standing genetic variation reduces the threshold population by approximately $65\%$. We observe that while standing genetic variation does not drive emergence of drug resistance when compared to de-novo aneuploids it does offer an advantage when compared with direct mutation.

@@

Even when evolutionary rescue occurs and leads to recurrence of the tumor, it may take a long time.
When the expected number of resistant lineages that avoid extinction is small the overall expected recurrence time can be estimated by adding two terms: the mean waiting time for evolutionary rescue--the appearance of a resistant lineage that avoid extinction (which we denote the mean evolutionary rescue time) and the expected time for proliferation of that lineage back to the original tumor size, $N$ (which we denote the mean proliferation time). However, when the expected number of resistant lineages is large the population of mutant cells behaves deterministically and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time. Of particular interest is the distribution of the evolutionary rescue time and recurrence time in the parameter regime  $\Delta_a T^*\ll1$ (partially tolerant aneuploids) for which we select  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ .

\paragraph{Evolutionary rescue time.}
In Appendix C we derive an approximation for $\tau_m$, the expected evolutionary rescue time without aneuploidy ($u=0$), and $\tau_a$, the expected rescue time with aneuploidy ($u>0$). \Cref{MeanTimeGrowthAneuploidyPlot,EvolutionaryRescueTimeComplete} shows the agreement between these approximations and simulation results for small, intermediate and large tumor sizes. Additionally, we note that $\tau_a$ has the following asymptotic behaviour:
\begin{equation}  \label{eq:AsymptoticTimeRules}
\tau_a \approx \begin{cases}
    \frac{1}{\abs{\Delta_w}}+\frac{1}{\abs{\Delta_a}} ,&
 N \ll N_a^* ,\\ 
  \left(v\lambda_wNp_m\right)^{-1}  ,&
  N \gg N_m^* .
  \end{cases}
\end{equation}
Interestingly, for small tumors the mean evolutionary rescue time is a function of the wildtype and aneuploid growth rates and independent of all of other model parameters including tumor size $N$ (\Cref{EvolutionaryRescueTimeComplete}). Increasing either the wildtype or aneuploid growth rates lead to an increase in the mean time. This can be interpreted that by increasing either growth rates the wildtype or aneuploid cells will last longer and will be able to produce a rescue mutation at latter times. Using the parameters that we have selected we observe that $\abs{\Delta_w}^{-1}=0.04$ and $\abs{\Delta_a}^{-1}=10^{-4}$ such that the mean evolutionary rescue time is determined by the aneuploid growth rate, i.e. $\tau_a\approx1/\abs{\Delta_a}$. For large tumors, we observe that the evolutionary rescue time is exponentially distributed with rate $v\lambda_wNp_m$ which is independent of parameters characterizing aneuploids or their production ($u$, $\lambda_a$, and $\Delta_a$). Increasing the mutation rate per mitosis $v\lambda_w$ leads to a faster appearance of a rescue mutation. Note that in the second case increasing the tumor size leads to smaller mean evolutionary rescue time as the wildtype is able to produce a rescue lineage faster.

When a fraction $f\approx 1.36\%$ of the initial cancer cell population is aneuploid we calculate the mean evolutionary rescue time in \Cref{meantimet2SGV} which we plot in \Cref{SGVEvolutionaryRescueTimeComplete}. We note that standing genetic variation reduces the mean evolutionary rescue time only for intermediary sized tumors but not for large or small tumors.
\paragraph{Recurrence time.}

In Appendix D we approximate the mean time for the population of mutant cancer cells to reach the initial population size $N$ which we denote as the recurrence time. \Cref{proliferationFigure} shows the agreement between our approximations and simulations. For values of $N$ smaller then $10^8$ the mean recurrence time can be approximated as the sum of the mean time for the first rescue mutation to appear and the mean time for the lineage generated by that mutation to reach size $N$. For $N>10^8$ the behavior of the mutant cancer cell population is deterministic and the mean recurrence time becomes constant. We note that $\tau_a^r$ has the following asymptotic behavior:
\begin{equation} \label{eq:AsymptoticTimeRulesRecurrence}
\tau_a^r \approx \begin{cases}
   \frac{1}{ \abs{\Delta_w}}+\frac{1}{ \abs{\Delta_a}}+\frac{\log p_mN}{\Delta_m} ,&
 N \ll N_a^* ,\\ 
  \frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_w}{v\lambda_w}  ,&
  N \gg N_m^* .
  \end{cases}
\end{equation}
In the first case, when tumor is small, the mean recurrence time grows logarithmically with tumor size $N$ and is the same order of magnitude as the mean evolutionary rescue time (\Cref{ProliferationTimeLarge}). Increasing the mutant growth rate $\Delta_m$ leads to shorter recurrence times while increasing the wildtype and aneuploid growth rates $\Delta_w$ and $\Delta_a$, retrospectively, increases the recurrence time. In the second case, when tumor is large, we observe that increasing either the mutant growth rare $\Delta_m$ or the mutation rate $v$ leads to a decrease in the time needed for the tumor to rebound to its initial size. Additionally, drugs which significantly increase the wildtype death rate $\mu_w$ and do not affect the division rate $\lambda_w$ delays cancer relapse. Consequently, patients which take such drugs require a longer period of monitoring to guarantee the effectiveness of the treatment.
\paragraph{Distribution of evolutionary rescue time with and without aneuploidy.}

@@

The recurrence can also be defined as the time necessary for the tumor size to reach a detection threshold $M$. We derive the mean recurrence time to detection size $M=10^7$ in Appendix D and plot in \Cref{RecurrencePlot} with simulations. We observe that  for small and intermediate sized tumors the effect of the detection size is negligible when compared to $\tau_a^r$. However, for large tumors the mean recurrence time to detection size $M$ decreases logarithmically with tumor size $N$.

@@

We find that aneuploidy can have a significant effect on evolutionary rescue as it reduces the threshold population size by at least one order of magnitude even when aneuploidy is tolerant (\Cref{rescue_prob}). Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in primary tumors in which the number of cells is large enough ($N\gg N_m^*\approx 4\times10^7$) for the appearance of resistant mutation directly in sensitive cells before these cells become extinct (\Cref{rescue_prob}).
However, aneuploidy can have a crucial role in evolutionary rescue of secondary tumors, in which the number of sensitive cells may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}. This can have an impact on the recurrence of cancer after the resection of the primary tumor  through secondary tumors which are too small to be detected and for which chemotherapy is employed to prevent  cancer relapse and are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}. This is reinforced by the fact that metastases have been shown to have higher chromosome missagregation rate two to three orders of magnitude higher then primary tumors~\citep{kimmel2023intra}.

@@

\section{Evolutionary rescue time}

@@

\section{Recurrence time}
We define the proliferation time $\tau_a^r$  to be the time it takes the population of mutant cancer cells to reach the initial tumor size $N$. The number of rescue lineages generated by the wildtype population is given by  \cref{eq:twosteplineage} (see \Cref{ExpectedNumberRescueLineages}):
\begin{equation*}
r_1\left(\infty\right)=\frac{uv\lambda_w\lambda_aNp_m}{\abs{\Delta_w}\abs{\Delta_a}}=\frac{N}{N_a^*},
\end{equation*}
where we ignore lineages created by direct mutation because we assumed $u\lambda_a > \max{(-\Delta_a, 1/T^*)}$, $N\ll N_m^*$ and used \Cref{eq:N_a}.

@@

\section{Distribution of evolutionary rescue time}
The probability that a successful mutant has been generated by time $t$ is given by:
\begin{align*}
P\left(rescue,t\right)&=P\left(T_a<t\right)\\
&=1-\exp\left\{-\left[r_1\left(t\right)+r_2\left(t\right)\right]\right\}\\
&=1-\exp\left\{-\left[\frac{uv\lambda_w\lambda_aNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)+ v\lambda_wNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right]\right\},
\end{align*}
where $T_a$ is the time at which the first mutant cell appears that will avoid extinction and which was defined in Appendix C.

As a result, the probability that a successful mutant has not been generated by time $t$ is:
\begin{equation}
1-P\left(rescue,t\right)=\exp\left\{-\left[\frac{uv\lambda_w\lambda_aNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)+ v\lambda_wNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right]\right\}.
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Appendix F: Distribution of recurrence time}
The probability distribution of the time that a lineage, consisting initially of a single cell, will reach size $N$ as time $t$ is given by the Gumbel distribution $\text{Gumb}_{max}\left(\frac{\log Np_m}{\Delta_m},\frac{1}{\Delta_m}\right)$~\citep{avanzini2019cancer} with probability density function:
\begin{align*}
G\left(t\right)=\e^{-p_mN\e^{-\Delta_mt}}.
\end{align*}
A mutant lineage initiated at time $s$, through aneuploidy, at rate $v\lambda_ap_ma_s$ reaches size $N$ before time $t$ with probability $G\left(t-s\right)$ where $s\leq t$. As a result, the number of successful mutant lineages which reach size $N$ by time $t$ can be approximated by inhomogeneous Poisson random variable with rate:
\begin{align*}
r\left(t\right)=v\lambda_ap_m\int_0^t a_sG\left(t-s\right)\,\d s
\end{align*}
where $a_s$ is aneuploid population size at time $s$ defined in \cref{aneuploidpopeq}. The proliferation time is defined as the first time the size of all lineages reaches $N$. When $N\ll\abs{\Delta_w}\abs{\Delta_a}/uv\lambda_w\lambda_ap_m$ there is at most a single mutant lineage that will survive and reach size $N$ (\Cref{ExpectedNumberRescueLineages}) and the probability that the size of that lineage has not reached $N$ by time $t$ is given by:
\begin{align}\nonumber
P\left(m_t\leq N\right)&=\exp\left[-r\left(t\right)\right]\\ \label{eqmNs}
&=\exp\left[-\frac{Nuv\lambda_w\lambda_ap_m}{\Delta_w-\Delta_a}\int_0^t\left[\e^{\Delta_ws}-\e^{\Delta_as}\right]\e^{-p_mN\e^{-\Delta_m\left(t-s\right)}}\,\d s\right].
\end{align}
When $N\gg\abs{\Delta_w}\abs{\Delta_a}/uv\lambda_w\lambda_ap_m$ the dynamics of the cancer cell populations is deterministic and approximated by the system of ODEs shown in \cref{detODE}. As a result, the size of the mutant cell population will always be below $N$ until time $\tau_a^r$ and will always be greater after:
\begin{align}\label{eqmNd}
P\left(m_t\leq N\right)=1-H\left(t-\tau_a^r\right),
\end{align}
where $H(x)$ is the Heaviside function:
\begin{equation*}
H\left(x\right) = \begin{cases}
    0 ,&
  x<0 ,\\ 
  1 ,&
  x\geq0 .
  \end{cases}
\end{equation*}
We plot \cref{eqmNs} and \cref{eqmNd} in Figure and compare with stochastic simulations and observe that our approximation are in agreement.

>>>>

We can categorize tumors by their size: small tumors with size $N<N_a^*$, intermediate tumors with size $N_a^*<N<N_m^*$, and large tumors with size $N>N_m^*$ .
For the parameter values in \Cref{table1} with $\lambda_a=0.0899,\mu_w=0.14, u=10^{-2}, v=10^{-7}$, we are in the partially resistant aneuploid case, and substituting in \cref{eq:N_a,eq:N_m}, we have $N_a^* \approx 4.4 \cdot 10^6$ and $N_m^* \approx 4 \cdot 10^7$.
Hence, we obtain the ratio $N^*_a/N^*_m \approx 0.11$ (\cref{eq:N_ratio}), that is, aneuploidy reduces the threshold tumor size by approximately 90\% 
Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4.19 \cdot 10^6$ cells for a wide variety of tumors as reported by \citet{avanzini2019cancer}.

@@

Therefore, the ratio of threshold tumor size with standing generation variation, $\tilde{N}^*_{a}$, is similar to the ratio with de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_w}$ is replaced with the aneuploidy cost, $s$, such that 
\begin{equation}\label{eq:Nsgv}
\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{s}{\abs{\Delta_w}}.
\end{equation}
Comparing these approximations to results of stochastic evolutionary simulations, we find that the approximations perform very well (\Cref{rescue_denovo}).  % TODO which approximations? unclear
Standing genetic variation will drive evolutionary rescue to the drug if wildtype growth rate $\Delta_w$ is very negative due to a stronger effect of the drug on sensitive cells, or if the aneuploidy cost $s$ is very small due to a low cost of aneuploidy in the pre-drug conditions.
In contrast, de-novo aneuploid cells will have a greater contribution to rescue if the aneuploidy cost, $s$, is large, the effect of the drug is weak ($\Delta_w$ is small), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
For example, with  $\lambda_w=0.1,\mu_w=0.14, u=10^{-2}, \tilde{u}=10^{-3}, s=7.3\cdot10^{-3}$, the ratio of the threshold tumor sizes for standing vs de-novo variation is $\tilde{N}^*_a/N^*_a \approx 1.825$, which means that de-novo genetic variation is  the main driver of evolutionary rescue.

Using \cref{eq:N_a,eq:N_m,eq:Nsgv}, we can find the ratio of threshold tumor size for rescue via standing genetic variation  to the threshold for direct mutation,
\begin{equation} \label{eq:N_sgv_ratio}
\frac{\tilde{N}^*_a}{N^*_m}=\frac{\tilde{N}^*_{a}}{N^*_{a}} \cdot \frac{N^*_a}{N^*_m} \approx \frac{s}{\abs{\Delta_w}}\begin{cases}
    \frac{\abs{\Delta_a}}{\tilde{u}\lambda_a} ,&
  \Delta_a T^* \ll -1 ,\\ 
  \frac{1}{\tilde{u}}\left(v \cdot \frac{\Delta_m}{\lambda_m}\right)^{1/2} ,&
  -1 \ll \Delta_a T^* \ll 1  ,\\ 
  v \frac{\Delta_m}{\lambda_m} \cdot \left(\tilde{u}\frac{\Delta_a}{\lambda_a}\right)^{-1}  ,&
   \Delta_a T^* \gg 1 .
  \end{cases}
\end{equation}
We observe that evolutionary rescue through direct mutation is more likely if the cost of aneuploidy $s$ is very large or the effect of the drug $\Delta_w$ is small.  In contrast, standing genetic variation will drive adaptation if the pre-treatment chromosome missegreagation rate $\tilde{u}$ is very large. The ratio does not depend on the rate of chromosome missegregation induced by the drug $u$. However, if the aneuploid growth rate $\Delta_a$ increases, then evolutionary rescue is driven by standing genetic variation. For the parameter values of  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, \tilde{u}=10^{-3}, v=10^{-7}$, we are in the partially resistant aneuploid case and obtain the ratio $\tilde{N}^*_a/N^*_m \approx 0.356$, which means that standing genetic variation reduces the threshold tumor size by approximately $65\%$. We observe that while standing genetic variation does not drive evolution of drug resistance when compared to de-novo aneuploidy, it does offer an advantage when compared with direct mutation.

@@

When evolutionary rescue occurs and leads to recurrence of the tumor, it may still take a long time.
When the expected number of resistant lineages that avoid extinction is small, the overall expected recurrence time can be estimated by adding two terms: the \emph{mean evolutionary rescue time}, which is the waiting time for appearance of a resistant lineage that avoids extinction (conditioned on such an even occurring in the first place), and the \emph{mean proliferation time}, which is the expected time for that lineage to reach the original tumor size, $N$.
However, when the expected number of resistant lineages is large, the population of mutant cells behaves deterministically and the mean recurrence time cannot be separated into the mean evolutionary rescue time and mean proliferation time. % TODO I dont like the word "behaves", please be specific. - what exactly is deterministic and why does it mean you cannot do the separation.
Of particular interest is the distribution of the evolutionary rescue time and recurrence time with partially tolerant aneuploid cells ($\Delta_a T^*\ll1$), for which we focus on the parameter values  $\lambda_w=0.1, \lambda_a=0.0899,\lambda_m=0.1,\mu_w=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$ .

\paragraph{Evolutionary rescue time.}
In appendix \ref{sec:appendix_rescue_time} we derive an approximation for $\tau_m$, the expected evolutionary rescue time without aneuploidy ($u=0$), and $\tau_a$, the expected rescue time with aneuploidy ($u>0$), both conditioned on evolutionay rescue occuring.
\Cref{EvolutionaryRescueTimeComplete,MeanTimeGrowthAneuploidyPlot} shows the agreement between these approximations and simulation results for small, intermediate and large tumor sizes.
Notably, the expected rescue time with aneuploidy, $\tau_a$, follows these expressions for very small and very large tumors,
\begin{equation}  \label{eq:AsymptoticTimeRules}
\tau_a \approx \begin{cases}
    \frac{1}{\abs{\Delta_w}}+\frac{1}{\abs{\Delta_a}} ,&
 N \ll N_a^* ,\\ 
  \left(v\lambda_wNp_m\right)^{-1}  ,&
  N \gg N_m^* .
  \end{cases}
\end{equation}

For small tumors, the mean rescue time is a function of the wildtype and aneuploid growth rates and independent the other model parameters, including tumor size (blue line in \Cref{EvolutionaryRescueTimeComplete}).
Increasing the wildtype or aneuploid growth rates leads to an increase in the mean rescue time, because the corresponding cells will survive for longer and will produce additional rescue mutations at latter times.
In our focus parameter regime, we have $\abs{\Delta_w}^{-1}=0.04$ and $\abs{\Delta_a}^{-1}=10^{-4}$, such that the mean rescue time is mainly determined by the aneuploid growth rate: $\tau_a \approx 10^4$.

For large tumors, the mean evolutionary rescue time is $v\lambda_wNp_m$ (\cref{eq:AsymptoticTimeRules}), % I removed the mention of exponential distribution as it was not explained and I dont see why you need it here
 which is independent of parameters characterizing aneuploid cells or their production ($u$, $\lambda_a$, and $\Delta_a$). 
 Increasing the per division mutation rate, $v\lambda_w$, leads to faster appearance of a rescue mutation and hence reduced mean rescue time. 
Finally, increasing the tumor size leads to shorter mean rescue time, as there are more wildtype cells that can mutate to become resistant.

When a fraction $f\approx 1.36\%$ of the initial cancer cell population is aneuploid we calculate the mean evolutionary rescue time in \cref{meantimet2SGV} which we plot in \Cref{SGVEvolutionaryRescueTimeComplete}. We note that standing genetic variation reduces the mean evolutionary rescue time only for intermediary sized tumors but not for large or small tumors.
\paragraph{Recurrence time.}

In appendix \ref{sec:appendix_recurrence_time} we approximate the mean time for the population of mutant cancer cells to reach the initial population size $N$ which we denote as the recurrence time. \Cref{proliferationFigure} shows the agreement between our approximations and simulations. For values of $N$ smaller then $10^8$ the mean recurrence time can be approximated as the sum of the mean time for the first rescue mutation to appear and the mean time for the lineage generated by that mutation to reach size $N$. For $N>10^8$ the behavior of the mutant cancer cell population is deterministic and the mean recurrence time becomes constant. We note that $\tau_a^r$ has the following asymptotic behavior:
\begin{equation} \label{eq:AsymptoticTimeRulesRecurrence}
\tau_a^r \approx \begin{cases}
   \frac{1}{ \abs{\Delta_w}}+\frac{1}{ \abs{\Delta_a}}+\frac{\log p_mN}{\Delta_m} ,&
 N \ll N_a^* ,\\ 
  \frac{1}{\Delta_m}\log\frac{\Delta_m-\Delta_w}{v\lambda_w}  ,&
  N \gg N_m^* .
  \end{cases}
\end{equation}
In the first case, when tumor is small, the mean recurrence time grows logarithmically with tumor size $N$ and is the same order of magnitude as the mean evolutionary rescue time (\Cref{ProliferationTimeLarge}). Increasing the mutant growth rate $\Delta_m$ leads to shorter recurrence times while increasing the wildtype and aneuploid growth rates $\Delta_w$ and $\Delta_a$, retrospectively, increases the recurrence time. In the second case, when tumor is large, we observe that increasing either the mutant growth rare $\Delta_m$ or the mutation rate $v$ leads to a decrease in the time needed for the tumor to rebound to its initial size. Additionally, drugs which significantly increase the wildtype death rate $\mu_w$ and do not affect the division rate $\lambda_w$ delays cancer relapse. Consequently, patients which take such drugs require a longer period of monitoring to guarantee the effectiveness of the treatment.
\paragraph{Distribution of evolutionary rescue time with and without aneuploidy.}

@@

The recurrence can also be defined as the time necessary for the tumor size to reach a detection threshold $M$. We derive the mean recurrence time to detection size $M=10^7$ in appendix \ref{sec:appendix_recurrence_time} and plot in \Cref{RecurrencePlot} with simulations. We observe that  for small and intermediate sized tumors the effect of the detection size is negligible when compared to $\tau_a^r$. However, for large tumors the mean recurrence time to detection size $M$ decreases logarithmically with tumor size $N$.

@@

We find that aneuploidy can have a significant effect on evolutionary rescue as it reduces the threshold tumor size by at least one order of magnitude even when aneuploidy is tolerant (\Cref{rescue_prob}). Interestingly, aneuploidy is unlikely to contribute to evolutionary rescue in primary tumors in which the number of cells is large enough ($N\gg N_m^*\approx 4\times10^7$) for the appearance of resistant mutation directly in sensitive cells before these cells become extinct (\Cref{rescue_prob}).
However, aneuploidy can have a crucial role in evolutionary rescue of secondary tumors, in which the number of sensitive cells may be below the detection threshold of $\sim10^7$  \citep{bozic2013evolutionary}. This can have an impact on the recurrence of cancer after the resection of the primary tumor  through secondary tumors which are too small to be detected and for which chemotherapy is employed to prevent  cancer relapse and are estimated to cause the majority of cancer-related deaths~\citep{chaffer2011perspective}. This is reinforced by the fact that metastases have been shown to have higher chromosome missagregation rate two to three orders of magnitude higher then primary tumors~\citep{kimmel2023intra}.

@@

\section{Evolutionary rescue time}\label{sec:appendix_rescue_time}

@@

\section{Recurrence time}\label{sec:appendix_recurrence_time}
We define the proliferation time $\tau_a^r$  to be the time it takes the population of mutant cancer cells to reach the initial tumor size $N$. The number of rescue lineages generated by the wildtype population is given by  \cref{eq:twosteplineage} (see \Cref{ExpectedNumberRescueLineages}):
\begin{equation*}
r_1\left(\infty\right)=\frac{uv\lambda_w\lambda_aNp_m}{\abs{\Delta_w}\abs{\Delta_a}}=\frac{N}{N_a^*},
\end{equation*}
where we ignore lineages created by direct mutation because we assumed $u\lambda_a > \max{(-\Delta_a, 1/T^*)}$, $N\ll N_m^*$ and used \Cref{eq:N_a}.

@@

\section{Distribution of evolutionary rescue time}
The probability that a successful mutant has been generated by time $t$ is given by:
\begin{align*}
P\left(rescue,t\right)&=P\left(T_a<t\right)\\
&=1-\exp\left\{-\left[r_1\left(t\right)+r_2\left(t\right)\right]\right\}\\
&=1-\exp\left\{-\left[\frac{uv\lambda_w\lambda_aNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)+ v\lambda_wNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right]\right\},
\end{align*}
where $T_a$ is the time at which the first mutant cell appears that will avoid extinction and which was defined in appendix \ref{sec:appendix_rescue_time}.

As a result, the probability that a successful mutant has not been generated by time $t$ is:
\begin{equation}
1-P\left(rescue,t\right)=\exp\left\{-\left[\frac{uv\lambda_w\lambda_aNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)+ v\lambda_wNp_m\frac{\e^{\Delta_w t}-1}{\Delta_w}\right]\right\}.
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Distribution of recurrence time}
The probability distribution of the time that a lineage, consisting initially of a single cell, will reach size $N$ as time $t$ is given by the Gumbel distribution $\text{Gumb}_{max}\left(\frac{\log Np_m}{\Delta_m},\frac{1}{\Delta_m}\right)$~\citep{avanzini2019cancer} with probability density function:
\begin{align*}
G\left(t\right)=\e^{-p_mN\e^{-\Delta_mt}}.
\end{align*}
A mutant lineage initiated at time $s$, through aneuploidy, at rate $v\lambda_ap_ma_s$ reaches size $N$ before time $t$ with probability $G\left(t-s\right)$ where $s\leq t$. As a result, the number of successful mutant lineages which reach size $N$ by time $t$ can be approximated by inhomogeneous Poisson random variable with rate:
\begin{align*}
r\left(t\right)=v\lambda_ap_m\int_0^t a_sG\left(t-s\right)\,\d s
\end{align*}
where $a_s$ is aneuploid population size at time $s$ defined in \cref{aneuploidpopeq}. The proliferation time is defined as the first time the size of all lineages reaches $N$. When $N\ll\abs{\Delta_w}\abs{\Delta_a}/uv\lambda_w\lambda_ap_m$ there is at most a single mutant lineage that will survive and reach size $N$ (\Cref{ExpectedNumberRescueLineages}) and the probability that the size of that lineage has not reached $N$ by time $t$ is given by:
\begin{align}\nonumber
P\left(m_t\leq N\right)&=\exp\left[-r\left(t\right)\right]\\ \label{eqmNs}
&=\exp\left[-\frac{Nuv\lambda_w\lambda_ap_m}{\Delta_w-\Delta_a}\int_0^t\left[\e^{\Delta_ws}-\e^{\Delta_as}\right]\e^{-p_mN\e^{-\Delta_m\left(t-s\right)}}\,\d s\right].
\end{align}
When $N\gg\abs{\Delta_w}\abs{\Delta_a}/uv\lambda_w\lambda_ap_m$ the dynamics of the cancer cell populations is deterministic and approximated by the system of ODEs shown in \cref{detODE}. As a result, the size of the mutant cell population will always be below $N$ until time $\tau_a^r$ and will always be greater after:
\begin{align}\label{eqmNd}
P\left(m_t\leq N\right)=1-H\left(t-\tau_a^r\right),
\end{align}
where $H(x)$ is the Heaviside function:
\begin{equation*}
H\left(x\right) = \begin{cases}
    0 ,&
  x<0 ,\\ 
  1 ,&
  x\geq0 .
  \end{cases}
\end{equation*}
We plot \cref{eqmNs} and \cref{eqmNd} in Figure and compare with stochastic simulations and observe that our approximation are in agreement.
